15yrsChange

### 2024-04-09 doc/15yrsChange.Rmd
final submission

Mammals whose population densities have been inflated due to human activities are also considered nuisance species [@murphy_invasive_2014; @clout_invasion_2008], and in United States (US) eastern deciduous forests, this includes white-tailed deer (*Odocoileus virginianus*). Their populations have dramatically increased in the past 50 years [@mcshea_science_2003], due in large part to global change [@cote_ecological_2004; @rooney_deer_2001; @brown_future_2000]. White-tailed deer preferentially browse on woody plant species in the seedling and sapling life stages, which reduces woody plant survival, growth and density [@mcgarvey_effects_2013a]. Chronic white-tailed deer overabundance can reduce understory plant diversity and decrease the abundance of traditionally dominant woody species (*e.g.*, oak, *Quercus* spp.; [@bugalho_effects_2013a; @habeck_communitylevel_2015] while giving an advantage to species they find less palatable, including both indigenous (*e.g.*, pawpaw, *Asimina tribola*; spicebush, *Lindera benzoin*) and non-indigenous (*e.g.*, garlic mustard, *Alliaria petiolata*) species (@averill_regional_2017). The strong impact of deer browsing, exacerbated by competition with dense deer-resistant understory vegetation and possibly climate change, results in dramatic reduction of canopy tree species in the understory [@mcgarvey_effects_2013a; @russell_interactions_2017; @miller_compounding_2019; @gorchov_differential_2021; @miller_overabundant_2023].

Increased canopy tree mortality in combination with reduced canopy species recruitment in the understory is likely to have profound impacts on the deciduous forests of eastern North America.  When global change pressures (*e.g.*, climate change, nuisance species) inhibit recruitment of canopy species, the stage is set for disturbances to push forest ecosystems over a tipping point, after which there is little chance that a forest will recover to its pre-disturbance state in the foreseeable future [@anderson-teixeira_altered_2013; @mcdowell_pervasive_2020; @turner_novel_2023a]. In the mid-Atlantic region of the eastern US, forests face a severe regeneration debt: there are insufficient juveniles of canopy tree species to replace the mature cohort of that species when they eventually die [@miller_compounding_2019; @miller_overabundant_2023]. The juvenile canopy trees that are most common in contemporary mid-Atlantic forests represent a more mesophytic set of species (e.g., maple, *Acer* spp.; American beech, *Fagus grandifolia*) than historically dominated much of the region (e.g., oak, *Quercus* spp.; hickory, *Carya* spp.; [@nowacki_climate_2015; @thompson_four_2013], but even these species associated with mesic conditions are often not present in sufficient abundance to replace the current canopy [@miller_compounding_2019].

@@

The study was conducted in the 25.6-hectare Forest Dynamics plot at the Smithsonian’s National Zoo and Conservation Biology Institute (SCBI) in Front Royal, Virginia (38$^\circ$ 53' 36.6''N, 78$^\circ$  08’ 43.4''W). The forest dynamics plot, which is located in the central Appalachian Mountains adjacent to Shenandoah National Park, is composed of mature secondary eastern deciduous forest. As part of the Appalachian Oak forest region, the canopy is dominated by tulip poplar (*Liriodendron tulipifera*), oak (*Quercus* spp.), and hickory (*Carya* spp.) and the understory is primarily composed of spicebush (*Lindera benzoin*), paw-paw (*Asimina triloba*), American hornbeam (*Carpinus caroliniana*), and witch hazel (*Hamamelis virginiana*). The land-use history of the site is varied, including periods of agricultural development and intensive logging, with dendrological data estimating canopy tree establishment around 1900 [@anonymous_smithsonian_2011, @bourg_initial_2013, @mcgarvey_effects_2013a]. The plot, which includes a four-hectare deer exclosure that has decreased the presence of deer since 1990, is divided into 640 quadrats, each measuring 20 x 20 meters. It is one of 78 sites in the Forest Global Earth Observatory (ForestGEO), a global network of forest dynamic plots that facilitates comparative forest ecology studies around the world [@anderson-teixeira_ctfsforestgeo_2015; @davies_forestgeo_2021].

@@

Though AWM increased throughout the plot AWM rates remained stable in a majority of individual species. The increasing trend of AWM was driven primarily by two species: white ash (*Fraxinus americana*) and black oak(*Quercus velutina*) (Figure `r mortality_by_genera`). *Fraxinus americana*, the host species for non-indigenous pest emerald ash borer (*Agrilus planipennis*), had the highest annual AWM of any species present on the plot, with an average flux of `r round(avg_fram,1)` Mg C ha^-1^ yr^-1^ and loss peaking at `r fram_peak` Mg C ha^-1^ yr^-1^ in 2020. *Quercus velutina* experienced the second-highest rate of AWM at a rate of `r format(round(avg_quve,1),nsmall = 1)` Mg C ha^-1^ yr^-1^. The average annual AWM of these two species (`r avg_fram_quve` Mg C ha^-1^ yr^-1^), was higher than that of all other species combined (`r avg_all_other` Mg C ha^-1^ yr^-1^). Moreover, both species exhibited significant increases in AWM since the start of the census, a trend also observed in three other species in the Oak genera—*Quercus rubra*, *Quercus prinus* and *Quercus alba*.

>>>>

Mammals whose population densities have been inflated due to human activities are also considered nuisance species [@murphy_invasive_2014; @clout_invasion_2008], and in United States (US) eastern deciduous forests, this includes white-tailed deer (*Odocoileus virginianus*). Their populations have dramatically increased in the past 50 years [@mcshea_science_2003], due in large part to global change [@cote_ecological_2004; @rooney_deer_2001; @brown_future_2000]. White-tailed deer preferentially browse on woody plant species in the seedling and sapling life stages, which reduces woody plant survival, growth and density [@mcgarvey_effects_2013b]. Chronic white-tailed deer overabundance can reduce understory plant diversity and decrease the abundance of traditionally dominant woody species (*e.g.*, oak, *Quercus* spp.; [@bugalho_effects_2013; @habeck_communitylevel_2015] while giving an advantage to species they find less palatable, including both indigenous (*e.g.*, pawpaw, *Asimina tribola*; spicebush, *Lindera benzoin*) and non-indigenous (*e.g.*, garlic mustard, *Alliaria petiolata*) species (@averill_regional_2017). The strong impact of deer browsing, exacerbated by competition with dense deer-resistant understory vegetation and possibly climate change, results in dramatic reduction of canopy tree species in the understory [@mcgarvey_effects_2013b; @russell_interactions_2017; @miller_compounding_2019; @gorchov_differential_2021; @miller_overabundant_2023].

Increased canopy tree mortality in combination with reduced canopy species recruitment in the understory is likely to have profound impacts on the deciduous forests of eastern North America.  When global change pressures (*e.g.*, climate change, nuisance species) inhibit recruitment of canopy species, the stage is set for disturbances to push forest ecosystems over a tipping point, after which there is little chance that a forest will recover to its pre-disturbance state in the foreseeable future [@anderson-teixeira_altered_2013; @mcdowell_pervasive_2020; @turner_novel_2023]. In the mid-Atlantic region of the eastern US, forests face a severe regeneration debt: there are insufficient juveniles of canopy tree species to replace the mature cohort of that species when they eventually die [@miller_compounding_2019; @miller_overabundant_2023]. The juvenile canopy trees that are most common in contemporary mid-Atlantic forests represent a more mesophytic set of species (e.g., maple, *Acer* spp.; American beech, *Fagus grandifolia*) than historically dominated much of the region (e.g., oak, *Quercus* spp.; hickory, *Carya* spp.; [@nowacki_climate_2015; @thompson_four_2013], but even these species associated with mesic conditions are often not present in sufficient abundance to replace the current canopy [@miller_compounding_2019].

@@

The study was conducted in the 25.6-hectare Forest Dynamics plot at the Smithsonian’s National Zoo and Conservation Biology Institute (SCBI) in Front Royal, Virginia (38$^\circ$ 53' 36.6''N, 78$^\circ$  08’ 43.4''W). The forest dynamics plot, which is located in the central Appalachian Mountains adjacent to Shenandoah National Park, is composed of mature secondary eastern deciduous forest. As part of the Appalachian Oak forest region, the canopy is dominated by tulip poplar (*Liriodendron tulipifera*), oak (*Quercus* spp.), and hickory (*Carya* spp.) and the understory is primarily composed of spicebush (*Lindera benzoin*), paw-paw (*Asimina triloba*), American hornbeam (*Carpinus caroliniana*), and witch hazel (*Hamamelis virginiana*). The land-use history of the site is varied, including periods of agricultural development and intensive logging, with dendrological data estimating canopy tree establishment around 1900 [@bourg_initial_2013b; @mcgarvey_effects_2013b]. The plot, which includes a four-hectare deer exclosure that has decreased the presence of deer since 1990, is divided into 640 quadrats, each measuring 20 x 20 meters. It is one of 78 sites in the Forest Global Earth Observatory (ForestGEO), a global network of forest dynamic plots that facilitates comparative forest ecology studies around the world [@anderson-teixeira_ctfsforestgeo_2015; @davies_forestgeo_2021].

@@

Though AWM increased throughout the plot AWM rates remained stable in a majority of individual species. The increasing trend of AWM was driven primarily by two species: white ash (*Fraxinus americana*) and black oak (*Quercus velutina*) (Figure `r mortality_by_genera`). *Fraxinus americana*, the host species for non-indigenous pest emerald ash borer (*Agrilus planipennis*), had the highest annual AWM of any species present on the plot, with an average flux of `r round(avg_fram,1)` Mg C ha^-1^ yr^-1^ and loss peaking at `r fram_peak` Mg C ha^-1^ yr^-1^ in 2020. *Quercus velutina* experienced the second-highest rate of AWM at a rate of `r format(round(avg_quve,1),nsmall = 1)` Mg C ha^-1^ yr^-1^. The average annual AWM of these two species (`r avg_fram_quve` Mg C ha^-1^ yr^-1^), was higher than that of all other species combined (`r avg_all_other` Mg C ha^-1^ yr^-1^). Moreover, both species exhibited significant increases in AWM since the start of the census, a trend also observed in three other species in the Oak genera—*Quercus rubra*, *Quercus prinus* and *Quercus alba*.
